---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GPSNorm

## Installation

⚠️ **Recommended: use the provided Devcontainer / Docker environment.**  
GPSNorm depends on **INLA**, which uses platform-specific compiled binaries.  
Different binary builds can lead not only to installation problems, but also to
unexpected model behavior or numerical differences across systems.  
For this reason, the container environment is the recommended way to run GPSNorm
and to reproduce results.

---

### Option 1 — VSCode Devcontainer (recommended)

Clone the repository:

    git clone https://github.com/tiny-quant/GPSNorm.git
    cd GPSNorm

Open the folder in VSCode and select

    Reopen in Container

The environment will be built automatically using

    GPSNorm/.devcontainer/docker-compose.yml

Requirements:

- Docker  
- VSCode  
- Dev Containers extension  

This setup ensures a consistent R / INLA environment.

---

### Option 2 — docker-compose

Start the same environment manually:

    git clone https://github.com/tiny-quant/GPSNorm.git
    cd GPSNorm/.devcontainer
    docker compose up -d

This launches the container used by the devcontainer configuration.

---

### Option 3 — Install from GitHub (advanced users)

Manual installation may produce different numerical results depending on
system libraries and INLA binaries. Use the container environment if
reproducibility is important.

    install.packages("remotes")
    remotes::install_github("tiny-quant/GPSNorm", dependencies = TRUE)

---

## Example

```{r example}
# devtools::load_all(".") # uncomment if opened in container.
library(GPSNorm)

path <- system.file(
    "extdata",
    "example_geomx_GSE186213.xlsx", package = "GPSNorm"
)

# Load and format data. 
data_covid <- GPSNorm::load_geomx_for_inla(
    path, 
    seg_covariates = c( 
        "AOISurfaceArea", 
        "AOINucleiCount",
        "DeduplicatedReads",
        "SequencingSaturation", 
        "AlignedReads", 
        "Mix"
    )
)

data_covid_spe <- GPSNorm::geomx_long_to_spe(
    data_covid, group_col = "Mix",
    extra_coldata = c(
        "AOISurfaceArea",
        "AOINucleiCount",
        "DeduplicatedReads",
        "SequencingSaturation", 
        "Slide_ID"
    )
)

# Run normalization. 
set.seed(1)

covid_fit <- GPSNorm::normalize_spe_inla_pac(
    data_covid_spe, 
    model_fun = GPSNorm::fit_inla_baseline_model,
    model_args = list(
        verbose = TRUE
    ), 
    group_col = "Mix_bin"
)
```

```{r}
# DE volcano plot.
covid_fit_bayes <- GPSNorm::de_volcano_bayes(
    covid_fit$spe, 
    covid_fit$fit$fit
)

covid_fit_bayes$plot 
```

```{r}
# DE posterior for a specific gene (SCGB1A1). 

de_posterior <- plot_de_posterior(
    covid_fit, 
    gene = "1524", 
    gene_name_col = "Gene_ID"
)

de_posterior$plot
```