---
title: "GPSNorm-Intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPSNorm-Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r}
library(GPSNorm)
```

Because of the sensitivity of INLA binary we can only guarantee exact reproducibility in the `./devcontainer/` environment. 

GPSNorm is designed to be applied to the standard excel files outputed by the Nanostring GeoMx Digital Spacial Profiler. 

```{r}
#| label: Load data. 

path <- system.file(
    "extdata",
    "example_geomx_GSE186213.xlsx", package = "GPSNorm"
)

# Load and format data. 
data_covid <- GPSNorm::load_geomx_for_inla(
    path, 
    seg_covariates = c( 
        "AOISurfaceArea", 
        "AOINucleiCount",
        "DeduplicatedReads",
        "SequencingSaturation", 
        "AlignedReads", 
        "Mix"
    )
)

data_covid_spe <- GPSNorm::geomx_long_to_spe(
    data_covid, group_col = "Mix",
    extra_coldata = c(
        "AOISurfaceArea",
        "AOINucleiCount",
        "DeduplicatedReads",
        "SequencingSaturation", 
        "Slide_ID"
    )
)
```

After covert the data to a `SpatialExperiment` object the normalization can be computed. 

```{r}
#| label: Apply normalization. 

set.seed(1)

covid_fit <- GPSNorm::normalize_spe_inla_pac(
    data_covid_spe, 
    model_fun = GPSNorm::fit_inla_baseline_model,
    model_args = list(
        verbose = TRUE
    ), 
    group_col = "Mix_bin"
)
```

A volcano plot to analyze the estimated differential expression can be created from the the fit object by calling:   

```{r}
#| label: DE volcano plot

covid_fit_bayes <- GPSNorm::de_volcano_bayes(
    covid_fit$spe, 
    covid_fit$fit$fit
)

covid_fit_bayes$plot 
```

The posterior of a single gene can be examined by calling: 

```{r}
#| label: DE posterior for SCGB1A1

de_posterior <- plot_de_posterior(
    covid_fit, 
    gene = "1524", 
    gene_name_col = "Gene_ID"
)

de_posterior$plot
```